# Course Project for Pactical Machine Learning

## Objective

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

The goal of your project is to predict the manner in which they did the exercise. This is the `classe` variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

## Set up the environment

```{r}
## install.packages("caret")
library(caret)
## install.packages("rpart")
library(rpart)
## install.packages("rpart.plot")
library(rpart.plot)
## install.packages("RColorBrewer")
library(RColorBrewer)
## install.packages("rattle")
library(rattle)
## install.packages("randomForest")
library(randomForest)
```

Set a seed to ensure reproducibility
```{r}
set.seed(12345)
```

## Load the data

```{r}
training <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
testing <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```


## Partioning the training set into two

Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```


## Cleaning the data

### Remove Near Zero Variance Variables

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)

myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
#To check the new N?? of observations
dim(myTraining)
```

### Remove the first variable so that it does not interfer with ML Algorithms (because it is just an ID):
```{r}
myTraining <- myTraining[c(-1)]
```

### Remove variables with too many NAs using a 60% threshold:
```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
		for(j in 1:length(trainingV3)) {
			if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
				trainingV3 <- trainingV3[ , -j] #Remove that column
			}	
		} 
	}
}
#To check the new N?? of observations
dim(trainingV3)

#Seting back to our set:
myTraining <- trainingV3
rm(trainingV3)
```

### Repeat on the testing data
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#To check the new N?? of observations
dim(myTesting)

#To check the new N?? of observations
dim(testing)

#Note: The last column - problem_id - which is not equal to training sets, was also "automagically" removed
#No need for this code:
#testing <- testing[-length(testing)]
```

In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
		if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
			class(testing[j]) <- class(myTraining[i])
		}      
	}      
}
#And to make sure Coertion really worked, simple smart ass technique:
testing <- rbind(myTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```

## Using ML algorithms for prediction: Decision Tree

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```

### Visualize the decision tree:
```{r}
fancyRpartPlot(modFitA1)
```

### Generate predictions:
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

### (Moment of truth) Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

## Using ML algorithms for prediction: Random Forests

```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

### Generate predictions:
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

### (Moment of truth) Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```

## Conclusion
Examining the accuracy of the two methods shows that the Random Forest predicted with higher accuracy.  Thus, we will apply it to make the predictions for submission.

### Generate predictions:
```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```

### Generate files for submission
```{r}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```
